---
title: "Homework 5"
author: "Kristi Gdanetz MacCready"
date: "02/05/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Box Sync/Schmidt_Lab/Bio201_teaching/W20_Protocols/Lab5/")
```

# Load packages
```{r Load packages, message=FALSE, warning=FALSE, include=FALSE, results='hide'}
library(tidyverse)
library(readxl)
library(broom)
library(cowplot)
set.seed(7)
```

Save any data frames generated to curated_data directory, save all plots to figures directory. Based on assumptions, clearly state which statistical test will be used and why. Include any statistics (p- or R-values) as comments after the tests.

# Question 1

Use the SCFA wide dataframe from Lab 5. Check assumptions, conduct statistical tests, make a plot, to answer the question, "Which supplement causes the greatest increase in acetate concentrations?"
```{r}
# import and filter data set (q1_df)
q1_df <- read_delim(file = "raw_data/scfa_wide.txt", 
                        delim = "\t", escape_double = FALSE, 
                        trim_ws = TRUE, na=c("NA"),
                        col_names = TRUE,
                        col_types = cols(participant_id = col_character(),
                                         semester = col_character(),
                                         supplement_consumed = col_character(),
                                         quantity_compliant = col_character(),
                                         frequency = col_character(),
                                         acetate_delta = col_double(),
                                         butyrate_delta = col_double(),
                                         propionate_delta = col_double())) %>%
  filter(semester != "Winter2015",
         quantity_compliant != "no") %>%
  select(-butyrate_delta, -propionate_delta)
```

### Check assumptions 
```{r}
# check sample sizes
q1_df %>%
  group_by(supplement_consumed) %>%
  summarise(sample_size = n())
```

```{r}
# check normality of each group 
s1 <- q1_df %>%
  filter(supplement_consumed == "Accessible") 
shapiro.test(s1$acetate_delta) #p-value = ?

s2 <- q1_df %>%
  filter(supplement_consumed == "Arabino") 
shapiro.test(s2$acetate_delta) #p-value = ?

s3 <- q1_df %>%
  filter(supplement_consumed == "BRMPS+Accessible")
shapiro.test(s3$acetate_delta) #p-value = ?

s4 <- q1_df %>%
  filter(supplement_consumed == "HiMaize") 
shapiro.test(s4$acetate_delta) #p-value = ?

s5 <- q1_df %>%
  filter(supplement_consumed == "HiMaize+BRMPS") 
shapiro.test(s5$acetate_delta) #p-value = ?

s6 <- q1_df %>%
  filter(supplement_consumed == "Inulin") 
shapiro.test(s6$acetate_delta) #p-value = ?
```

```{r}
# check variances 
bartlett.test(acetate_delta ~ supplement_consumed, data = q1_df)
```

### Test 
```{r}
# anova
aov_q1 <- aov(acetate_delta ~ supplement_consumed, data = q1_df)
summary(aov_q1)

# kruskall-wallace
kruskal.test(acetate_delta ~ supplement_consumed, data = q1_df)
```

### Plot
```{r}
# make and save plot (q1_plot)
q1_plot <- q1_df %>%
  ggplot(aes(x = supplement_consumed,
             y = acetate_delta,
             color = supplement_consumed)) +
  geom_hline(yintercept = 0,
             linetype = "dashed") + 
  geom_violin() + 
  geom_jitter() + 
  xlab(NULL) + 
  ylab("Acetate mmol/kg (wk3 - wk1)") + 
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) 

save_plot(q1_plot, nrow = 1, ncol = 1, base_height = 4, base_width = 8,
          path = "~/Box Sync/Schmidt_Lab/Bio201_teaching/W20_Protocols/Lab5/figures/q1_plot.pdf")
```

What are your conclusions regarding the reseach question? 

<type answer here>


# Question 2

Use the SCFA wide dataframe from Lab 5. Check assumptions, conduct statistical tests, make a plot, to answer the question, "Which supplement causes the greatest increase in propionate concentrations?"
### Import & filter
```{r}
# import and filter data set (q2_df)
q2_df <- read_delim(file = "raw_data/scfa_wide.txt", 
                        delim = "\t", escape_double = FALSE, 
                        trim_ws = TRUE, na=c("NA"),
                        col_names = TRUE,
                        col_types = cols(participant_id = col_character(),
                                         semester = col_character(),
                                         supplement_consumed = col_character(),
                                         quantity_compliant = col_character(),
                                         frequency = col_character(),
                                         acetate_delta = col_double(),
                                         butyrate_delta = col_double(),
                                         propionate_delta = col_double())) %>%
  filter(semester != "Winter2015",
         quantity_compliant != "no") %>%
  select(-butyrate_delta, -acetate_delta)
```

### Check assumptions
```{r}
# check sample sizes
q2_df %>%
  group_by(supplement_consumed) %>%
  summarise(sample_size = n())
```

```{r}
# check normality of each group 
s1 <- q2_df %>%
  filter(supplement_consumed == "Accessible") 
shapiro.test(s1$propionate_delta) #p-value = ?

s2 <- q2_df %>%
  filter(supplement_consumed == "Arabino") 
shapiro.test(s2$propionate_delta) #p-value = ?

s3 <- q2_df %>%
  filter(supplement_consumed == "BRMPS+Accessible")
shapiro.test(s3$propionate_delta) #p-value = ?

s4 <- q2_df %>%
  filter(supplement_consumed == "HiMaize") 
shapiro.test(s4$propionate_delta) #p-value = ?

s5 <- q2_df %>%
  filter(supplement_consumed == "HiMaize+BRMPS") 
shapiro.test(s5$propionate_delta) #p-value = ?

s6 <- q2_df %>%
  filter(supplement_consumed == "Inulin") 
shapiro.test(s6$propionate_delta) #p-value = ?
```

```{r}
# check variances 
bartlett.test(propionate_delta ~ supplement_consumed, data = q2_df)
# enter arguments with the following formula: continuous ~ categorical 
```

### Statistical test 
```{r}
# anova
aov_q2 <- aov(propionate_delta ~ supplement_consumed, data = q2_df)
summary(aov_q2)

# kruskall-wallace
kruskal.test(propionate_delta ~ supplement_consumed, data = q2_df)
```

### Plot 
```{r}
# make and save plot (q2_plot)
q2_plot <- q2_df %>%
  ggplot(aes(x = supplement_consumed,
             y = propionate_delta,
             color = supplement_consumed)) +
  geom_hline(yintercept = 0,
             linetype = "dashed") + 
  geom_violin() + 
  geom_jitter() + 
  xlab(NULL) + 
  ylab("Propionate mmol/kg (wk3 - wk1)") + 
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) 

save_plot(q2_plot, nrow = 1, ncol = 1, base_height = 4, base_width = 8,
          path = "~/Box Sync/Schmidt_Lab/Bio201_teaching/W20_Protocols/Lab5/figures/q2_plot.pdf")
```

What are your conclusions regarding the reseach question?

<type answer here>


# Question 3

Use the SCFA wide dataframe from Lab 5. Subset and filter the data frame, check assumptions, and choose the appropriate statistical tests, and create a plot to answer the following research question: "Does the proportion of acetate reponses differ between individuals who consumed BRMPS potato (BRMPS) or LOODAT potato supplements?"
### Import & filter 
```{r}
q3_df <- read_delim(file = "raw_data/scfa_wide.txt", 
                        delim = "\t", escape_double = FALSE, 
                        trim_ws = TRUE, na=c("NA"),
                        col_names = TRUE,
                        col_types = cols(participant_id = col_character(),
                                         semester = col_character(),
                                         supplement_consumed = col_character(),
                                         quantity_compliant = col_character(),
                                         frequency = col_character(),
                                         acetate_delta = col_double(),
                                         butyrate_delta = col_double(),
                                         propionate_delta = col_double())) %>%
  filter(supplement_consumed == "LOODAT" | supplement_consumed == "BRMPS",
         semester != "Winter2015",
         quantity_compliant != "no") %>%
  select(-butyrate_delta, -propionate_delta) %>%
  add_column(acetate_status = "-") %>%
  drop_na()

q3_df$acetate_status[q3_df$acetate_delta > 0] <- "positive"
q3_df$acetate_status[q3_df$acetate_delta < 0] <- "negative"
```

### Check assumptions
* Random samples (this is met because supplements are randomly assigned)
* Normally distributed data (verify in same method as a t-test)
* Independent samples (the two groups in our question are not linked in any way)
```{r}
# normality check in positive group
p1 <- q3_df %>%
  filter(acetate_status == "positive") 
shapiro.test(p1$acetate_delta) #p-value = ?

# normality check in negative group
n1 <- q3_df %>%
  filter(acetate_status == "negative") 
shapiro.test(n1$acetate_delta) #p-value = ?
```

### Proportion test
```{r}
# create count table 
q3_tab <- with(q3_df, table(supplement_consumed, acetate_status))

# conduct statistical test
prop.test(q3_tab, 
          alternative = "two.sided",
          correct = FALSE)
```

### Make table 

| | Increased acetate | Total participants | Percent increase |
|:-----:|:-----:|:-----:|:-----:|
| BRMPS | N | Y | Z% |
| LOODAT | n | y | z% | 


What are your conclusions regarding the reseach question?

<type answer here>


# Question 4

Use the SCFA wide dataframe from Lab 5. Subset and filter the data frame, check assumptions, and choose the appropriate statistical tests, and create a plot to answer the following research question: "Does the proportion of propionate reponses differ between individuals who consumed BRMPS potato (BRMPS) or LOODAT potato supplements?"
### Import & filter
```{r}
# import and filter data set (q4_df)
q4_df <- read_delim(file = "raw_data/scfa_wide.txt", 
                        delim = "\t", escape_double = FALSE, 
                        trim_ws = TRUE, na=c("NA"),
                        col_names = TRUE,
                        col_types = cols(participant_id = col_character(),
                                         semester = col_character(),
                                         supplement_consumed = col_character(),
                                         quantity_compliant = col_character(),
                                         frequency = col_character(),
                                         acetate_delta = col_double(),
                                         butyrate_delta = col_double(),
                                         propionate_delta = col_double())) %>%
  filter(supplement_consumed == "LOODAT" | supplement_consumed == "BRMPS",
         semester != "Winter2015",
         quantity_compliant != "no") %>%
  select(-butyrate_delta, -acetate_delta) %>%
  add_column(propionate_status = "-") %>%
  drop_na()

q4_df$propionate_status[q4_df$propionate_delta > 0] <- "positive"
q4_df$propionate_status[q4_df$propionate_delta < 0] <- "negative"
```

### Check assumptions
* Random samples (this is met because supplements are randomly assigned)
* Normally distributed data (verify in same method as a t-test)
* Independent samples (the two groups in our question are not linked in any way)
```{r}
# normality check in positive group
p2 <- q4_df %>%
  filter(propionate_status == "positive") 
shapiro.test(p2$propionate_delta) #p-value = ?

# normality check in negative group
n2 <- q4_df %>%
  filter(propionate_status == "negative") 
shapiro.test(n2$propionate_delta) #p-value = ?
```

### Proportion test
```{r}
# create count table 
q4_tab <- with(q4_df, table(supplement_consumed, propionate_status))

# conduct statistical test
prop.test(q4_tab, 
          alternative = "two.sided",
          correct = FALSE)
```

### Make table 

| | Increased propionate | Total participants | Percent increase |
|:-----:|:-----:|:-----:|:-----:|
| BRMPS | N | Y | Z% |
| LOODAT | n | y | z% | 


What are your conclusions regarding the reseach question?

<type answer here>


-----
end